---
title: 'Lab 5: Simulating discrete probability models'
author: 'Justin Pelak'
output:
  word_document: default
  pdf_document: default
date: "Mar 24, 2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1: Roulette wheel simulation
A roulette wheel has 38 slots of which 18 are red, 18 are black, and 2 are green. If a ball spun on to the wheel stops on the color a player bets, the player wins. Consider a player betting on red. Winning streaks follow a Geometric(*p* = 20/38) distribution in which we look for the number of red spins in a row until the first black or green. Use the derivation of the Geometric distribution from the Bernoulli distribution to simulate the game. Namely, generate Bernoulli(*p* = 20/38) random variates (0 = red; 1 = black or green) until a black or green occurs. 

### Code set-up
A `while` loop allows us to count the number of spins until a loss.  If we use indicator variable `lose` to note a win (1) or loss (0), the syntax is "while we have not lost (i.e., `lose==0`), keep spinning."  Once you win, the while loop ends and the variable `streak` has counted the number of spins.  Try running a few times.
```{r spins}
streak = 0
lose = 0
p = 20/38
while(lose==0){
  lose = (runif(1) < p) # generate Bernoulli with probability p
  streak = streak + 1 # tally streak
}
streak
```

### The problem
The code chunk above performs the experiment once: spin the roulette wheel until you lose and record the number of spins.  Simulate 1000 experiments.  As usual, do this by wrapping the code chunk above within a for-loop and storing the number of spins `streak` in a vector.
```{r rand, echo=TRUE}
# [Place code here]
```

### Report the following:

* Histogram of the win streak length.  Note that this is a discrete distribution so should place histogram bars at discrete values {0, 1, 2, ...}.  This may be done with the `breaks` option within `hist`.  If your storage variable is called `winstreak`:

`hist(winstreak, br=seq(min(winstreak)-0.5, max(winstreak+0.5)), main="")`

* Average length of the win streak. *[Answer here]*

* Standard deviation of the winning streak lengths. *[Answer here]*

* Compare the empirical average and standard deviation in the previous two bullets to the true values from the Geometric(*p* = 20/38) distribution.

*[Answer here]*

* Longest winning streak. *[Answer here]*

\vfill

## Task 2: Simulating negative binomial distributions

In this task, we will compare two different algorithms for simulating from a negative binomial distribution.

### Problem (a)
Recall that a negative binomial random variable *NB(r, p)* is the sum of *r Geometric(p)* random variables.  Use the algorithm from Task 1 to simulate 1000 *NB(10, 0.6)* random variates.

### Code set-up
Note that we merely need to wrap the core code from Task 1 within a for-loop.  Here is the core of the code chunk, where we are thinking of a for-loop over a variable `sims` to replicate the single negative binomial draw.  Note that this code chunk will not run since the for-loop over `sims` is not coded, thus the `eval=FALSE` option.  **Note that this code has the `eval=FALSE` option just to present the code without output.  Your code will not use this option.**

```{r codecore, eval=FALSE}
for(nbsims in 1:r){
  # for-loop allows us to simulate until r successes;
  # in this problem, r=10 and p=0.6
  tossnum = 0
  success = 0
  while(success==0){
    success = (runif(1)<p)
    tossnum = tossnum + 1
  }
  nbvar1[sims] = nbvar1[sims] + tossnum
}
```


### Problem (b)
The negative binomial pmf induces the following recursion relation.  If $X \sim B(r,\ p)$, then
\[ P(X = i+1) = \frac{i(1-p)}{(i+1-r)} \cdot P(X=i).\]
Use this recursion relation to generate 1000 $NB(10,\ 0.6)$ random variates.

### Code set-up
Below is binomial.R, the binomial simulator used in the video lectures and found also on the class Blackboard site.  

```{r binomial, echo=TRUE}
simnum = 1000
p = 0.6; r = 10  # for point of comparison with the negative binomial, we will use r here
y=0
for(sims in 1:simnum){
	pmf=(1-p)^r; cdf=pmf; # pmf and cdf
	j=0;
	u=runif(1) # uniform random variate
	# find Binomial variate
	while(u >= cdf){
		pmf=((r-j)/(j+1))*(p/(1-p))*pmf # recursion relation
		cdf=cdf + pmf # compute cdf
		j=j+1
	}
	y[sims]=j
}
```

This binomial simulator may be applied directly after changing just three lines: 

* $j=r$

* the recursion relation formula

* `pmf = p^r`

### Report the following for each of the simulations in problems (a) and (b)

* Histogram of the variates

* Mean and standard deviation of the simulated variates

* Run time: compare computing speed between the two algorithms.  In R, can wrap your algorithm or sequence of operations as follows to time your code.
```{r timewrapper}
x = proc.time()
# [the code you want to time here]
timer = proc.time() - x
algtime = timer[3]  # algtime will store the algorithm run time in seconds
```

### Questions

* How do the histograms compare?

*[Answer here]*

* How do the mean and standard deviation from the simulations compare to the true mean and standard deviation of a $NB(0.6,\ 10)$ distribution?

*[Answer here]*

* How do the computing times compare?  Which algorithm is faster?

*[Answer here]*

* "Simulation flops": Which simulator do you think uses more uniform random numbers (call to the `runif()` function)?  Why?

*[Answer here]*