---
title: "RCodingAssignment3"
author: "Justin Pelak"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
If 10% of people who are given a certain drug experience Dizziness, find these probabilities using R code for a sample of 25 people who take the drug.

### A. Exactly 3 people will become dizzy
```{r "1a"}
n=25
p=0.10
k=3
dbinom(k,n,p)
```

### B. At least 9 people will become dizzy.
```{r "1b"}
k=9
pbinom(k-1,n,p,lower.tail=F)
```

### C. Less than 7 people will become dizzy
```{r "1c"}
k=6
pbinom(k,n,p)
```

## Question 2
There were ten green bottles sitting on the wall. The probability of a green bottle accidentally falling is 0.95. What is the probability that fewer than 8 of the green bottles accidentally fall?

```{r "q2"}
n=10
p=0.95
k=7
pbinom(k,n,p)
```

## Question 3
The LMB Company manufactures tires. They claim that only .007 of LMB tires are defective. What is the probability of finding 2 defective tires in a random sample of 50 LMB tires?

```{r "q3"}
n=50
p=0.007
k=2
dbinom(k,n,p)
```
